Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states.

### 2023-06-09 inst/manuscript/Response.md

## Editor and Reviewer comments

### Reviewer #1

> 1.  Please consider reference style: Number the references (numbers in square brackets) in the list in the order in which they appear in the text.

{\>\>Owen: Rainer - please correct and then make a response here.\<\<} {\>\>Rainer: Will do this at the end\<\<} \*

@@

> The authors in section 3.3 claim several sentences that could confuse readers. 2.1 `one discovers how the steady state of the system responds to the environmental conditions`. There are several concerns that the authors should clarify:
>
> a)  `The steady state`. Is there a unique steady state? Is there a reported contribution concerning a formal analysis of the model in the sense of stability (stable state existence and uniqueness)? Maybe, authors should refer as `a system's stable state (if it exists)`.

**Response:** Thank you for raising this important issue, which we had overlooked. A number of the reviewer comments below are closely related, hence we sometimes refer to this response and do so using the label "Response R1". In order to clarify about the existence and uniqueness of steady states we have added the text `When one wishes to be able to make conclusions about how the steady state of the system is affected by the environmental driver, it is very important to note that the *final state* (found by the simulation) is not guaranteed to be a *steady state*. The user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper @Limberger2022, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.`

**Response:** In addition, we notice that we sometimes used the term "stable state" and sometimes "steady state". We now use only the term "steady state". Furthermore, we now use the term "final state" to refer to the state of the system at the end of a simulation.

> `b)`The steady state of the system responds\`. It should be noted that the steady state does not respond to the environment. Maybe, the authors refer to the numerical result of a steady state as a function of parameter and initial condition (if a steady state is not unique).

@@

**Response:** Thank you. We have edited the text so that we do not state that the steady state responds to the environment.

> 2.2 `finding steady states that correspond to values of one environmental driver`. The software does not find a steady state. Maybe the authors refer to the software running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (oxygen diffusivity). Then, the numerical evaluation of steady states could be related to environmental conditions in real systems.

**Response:** We envisage and hope that Response R1 sufficiently clarifies steady state finding.

@@

> c)  `in order to check if results were sensitive to choice of method`. This sentence needs to be clarified. There is no comparison between numerical evaluations in any sense. The numerical simulations or simulation results are not sensitive to choice or method. Indeed, the numerical solution of the dynamical model is sensitive to a set of parameters and initial conditions. Maybe the authors mean the following: to evaluate the performance of the numerical parametric sensitivity analysis of the dynamical model concerning one parametric variation (oxygen diffusivity) under two schemes.2.7 It is not clear why there are two `methods` for numerical parametric sensitivity analysis. The authors should include the arguments of this.

**Response:** We removed this text and now simply state that having two methods allows the comparison of their results.

> 2.8 Please consider rewrite the section 3.3 (see previous 2.1 - 2.7 points) in the sense of Parametric sensitivity analysis, Open loop dynamical behavior, Equilibrium points (see Khalil, H. K. (2002). Nonlinear Systems. Prentice Hall, New Jersey.)

**Response:** We have the opinion that the with the revisions already made and detailed above, and those below, that the terminology is now more in line with this text, and will be clearly understandable by the target audience (theoretical ecologists) even when the section is not more substatially rewritten.

@@

> Line 119 `The general approach used to find the stable state of the system`

Changed to `The general approach to finding the numerical value of a system's stable state (if it exists)`

> Line ...
>
> 4.  Please, consider English Editing Service or Proofread Your Paper.

The paper has been proofread and edited.

{\>\>Owen: Rainer - please would you see to this?\<\<}

@@

> 5.  Please, consider include two or three references of SoftwareX Journal.

{\>\>Owen: Rainer - please would you see to this?\<\<}

------------------------------------------------------------------------

------------------------------------------------------------------------

### 2023-06-09 inst/manuscript/manuscript_microxanox.Rmd
Further work on the revision

---
title: "Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states."
author:
- name: Rainer M Krug
  email: Rainer.Krug@uzh.ch;  Rainer@krugs.de
  affiliation: University of Zürich
  footnote: Corresponding Author
- name: Owen L. Petchey
  email: Owen.Petchey@ieu.uzh.ch
  affiliation: University of Zürich
address:
- code: University of Zürich
  address: Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse
    190, 8057 Zurich
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
keywords: 
  - reproducibility
  - regime shift 
  - steady state
  - ordinary differential equations
abstract: |
  $Microxanox$ is an R package to simulate a three functional group ecosystem (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range, to find steady states.  The model can be implemented with different numbers of species per functional group. The package is constructed in such a way that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated. Furthermore, the package framework and code should serve as a useful starting point for making simulation models of other types of ecosystems.
journal: "SoftwareX"
date: "`r Sys.Date()`"

@@

<!-- # \textcolor{red}{TODO} -->
<!-- - [x] ~~**\textcolor{green}{update links}**~~ -->
<!-- - [x] ~~**\textcolor{green}{update cross references}**~~ -->
<!-- - [x] ~~**\textcolor{green}{add abstract}**~~ -->
<!-- - [x] ~~**\textcolor{green}{check keywords}**~~ -->
<!-- - [x] ~~**\textcolor{green}{add more references}**~~ -->
<!-- - [x] ~~**\textcolor{green}{expand on Impact section}**~~ -->
<!-- - [x] ~~**\textcolor{green}{expand on Conclusion section}**~~ -->
\pagebreak

# Required Metadata
## Current code version
Ancillary data table required for subversion of the codebase.

| **Nr.** | **Code metadata description**                                | **Please fill in this column**                                                 |  
| :-----: | :----------------------------------------------------------- | :----------------------------------------------------------------------------- |  
|   C1    | Current code version                                         | v0.9.1                                                                         |  
|   C2    | Permanent links to code/repository used for this code version | [https://github.com/UZH-PEG/microxanox](https://github.com/UZH-PEG/microxanox) | 
|         |                                                              | [10.5281/zenodo.7148667](https://zenodo.org/record/7148667)                    |
|   C3    | Code Ocean compute capsule                                   |                                                                                |  
|   C4    | Legal Code License                                           | MIT                                                                            |  
|   C5    | Code versioning system used                                  | [git](https://git-scm.com)                                                                            |  
|   C6    | Software code languages, tools, and services used            | [R](https://cran.r-project.org/index.html)                                     |  
|   C7    | Compilation requirements, operating environments             | [R (>= 4.1.0)](https://cran.r-project.org/index.html)                          |  
|         |                                                              | magrittr                                                                       |  
|         |                                                              | tibble                                                                         |  
|         |                                                              | ggplot2                                                                        |  
|         |                                                              | patchwork                                                                      |  
|         |                                                              | grDevices                                                                      |  
|         |                                                              | stats                                                                          |  
|         |                                                              | mgcv                                                                           |  
|         |                                                              | deSolve                                                                        |  
|         |                                                              | dplyr                                                                          |  
|         |                                                              | tidyr                                                                          |  
|         |                                                              | stringr                                                                        |  
|         |                                                              | multidplyr                                                                     |  
|   C8    | If available Link to developer documentation/manual          | [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html) |  
|   C9    | Support email for questions                                  | Rainer.Krug@uzh.ch; Owen.Petchey@ieu.uzh                                       |

@@

Mathematical models play a key role in the development of ecosystem models and the understanding about how and why ecosystems change when the environment changes [@Binzer2016a; @Chaparro-Pedraza2021; @Vasseur2005]. They are also important for developing hypotheses to test in empirical studies. One area of ecology in which models have been influential is the understanding of their response to a gradual change in an environmental driver [@Scheffer2001]. An environmental driver is an environmental condition that affects an ecosystem. Still, it is assumed not to be affected by the ecosystem, such as the rate of nutrient input into a lake.

It is conceivable that an ecosystem state, such as the total biomass of a particular type of bacteria, may remain unchanged when an environmental driver changes. It is also possible that the ecosystem state changes gradually. It is also possible that the ecosystem state changes abruptly to a new state that is difficult to recover from [@Scheffer2001]. This possibility for abrupt, perhaps catastrophic changes that are difficult to reverse causes considerable concern [@Collins2021; @Northrop2021; @Vandermeer2019].

An example where a gradual change of an environmental variable causes an abrupt change in the system is the switch from an aerobic (oxygen is available for metabolism) to an anaerobic (oxygen generally unavailable) state in a microbial ecosystem. This system has been investigated by @Bush2017 in a simulation study of a mathematical model. Three types of microbes occur in the model: cyanobacteria (CB) dominating the oxic state, and two types of sulfur bacteria that dominate the anoxic state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria (PB)). The model shows that gradual change in the rate at which oxygen could diffuse into the ecosystem (termed oxygen diffusitivity) could cause catastrophic changes in the ecosystem state that would be difficult to reverse.

One feature of the study by @Bush2017 was limited biodiversity. Specifically, there was no biodiversity within each of the three types of bacteria. This leaves the question of if and how bacteria's biodiversity within these types (i.e., functional groups) affects the ecosystem response to environmental change. This limitation is not specific to the study of @Bush2017. There are few if any studies of the effects of biodiversity on abrupt transitions between ecosystem states.

@@

# Software description
<!-- > {>>Describe the software in as much as is necessary to establish a vocabulary needed to explain its impact.<<} -->

@@

To make the simulation run with multiple species per functional group, we expressed different species characteristics in the elements of vectors and matrices. We also coded the ordinary differential equations to include the vectors and matrices and used matrix mathematics. In this way, we made it possible to run simulations with different numbers of species without having to change the underlying code.

@@

## Software architecture
<!-- > {>>Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.<<} -->

The framework used when writing this package aimed to maximise simplicity for the user and to make it straightforward to reproduce results (see the supplement [@Petchey2022] to @Limberger2022 for an example of how this is used). As such, all the parameters needed to run a simulation or find a steady state (i.e. the final state of the ecosystem) are contained in a single object (which can easily be created using included functions). This parameter object is given to a function that runs the simulations and returns the results. The returned results object is identical to the parameter object but with an additional slot named `results`, which contains the simulation results. Thus the returned results object contains the simulation conditions (parameters) as well as the results, and can be used to run the simulation again. This promotes reproducibility and makes incremental changes of individual parameters with a consecutive re-running of the simulations straightforward.

In the following sections we describe how to use the package to run one simulation and to find steady states across an environmental gradient.

@@

## Finding Steady States

The general approach to finding the numerical value of a system's steady state (if it exists) with a specific parameter set is to run the simulation for a long time and record the final state. When one does this across a range of environmental conditions, one discovers how the steady state of the system changes with the environmental conditions. The package contains functionality for finding steady states that correspond to values of one environmental driver, namely the value of oxygen diffusivity.

Two numerical strategies for finding steady states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow comparison of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study and was used to obtain the results in figures 3 and 4 of that article.

The second method runs two simulations, one with step-wise and slowly temporally *increasing* oxygen diffusivity, and the other with step-wise and slowly *decreasing* oxygen diffusivity. (Put another way, to explore the sensitivity of steady states of the dynamic model under oxygen diffusivity variation, subset values of this parameter are numerically evaluated. Two value vectors of oxygen diffusivity are presented step-wise, gradually increasing and decreasing.) During this temporal environmental change, the state of the system is recorded just before change to a new oxygen diffusivity (we term this the *Temporal method*).

The package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique, and indeed if a steady state has been achieved.

@@

From the results returned, summary measures about how the ecosystem steady states changes with environmental change can be extracted. The function `get_stability_measures()` returns quantities such as the amount of environmental change required for the system to abruptly change to a different state.

@@

<!-- > {>>This is the main section of the article and the reviewers weight the description here appropriately -->
<!-- Indicate in what way new research questions can be pursued as a result of the software (if any). -->
<!-- Indicate in what way, and to what extent, the pursuit of existing research questions is improved (if so). -->
<!-- Indicate in what way the software has changed the daily practice of its users (if so). -->
<!-- Indicate how widespread the use of the software is within and outside the intended user group. -->
<!-- Indicate in what way the software is used in commercial settings and/or how it led to the creation of spin-off companies (if so).<<} -->

The open source implementation and extension of the model used in @Bush2017 provides the means of reproducing the results published while at the same time provides the means of doing unique, innovative, and important investigations of how ecosystems change as the environmental changes, and how biodiversity may modulate this.

The design of the package code and functionality is with reproducibility in mind: the combination of all parameters being in a single parameter object as well as the return of the simulation as a result object which inherits from the parameter object provides a relatively easy to use framework to implement reproducible experiments.

Here we evidence the impact of the *microxanox* package by describing three use cases and then by describing how the package can be a starting point for models of other ecosystems. The first two use cases are described in detail (including the code for reproducing them) in the [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html) and the *Partial Reproduction* vignettes. The third is taken from @Limberger2022 and @Petchey2022.

@@

The study of @Bush2017 includes simulations of the effect of oxygen diffusivity (an environmental driver, in the sense that it affects the ecosystem but is not affected by it) on the ecosystem state (oxic or anoxic). The `microxonox` package contains functionality to make a specific temporal pattern of change in the oxygen diffusivity. As well as allowing individual simulations during which oxygen diffusivity varies, this functionality forms the basis of the temporal method for finding steady states.

An example of this functionality is given in the *Partial Reproduction* vignette, which we briefly show here (Figure \@ref(fig:uc1)). The example is composed of a single simulation, at the beginning of which the system is in the oxic state with high abundance of cyanobacteria. Oyxgen diffusivity is then slowly decreased and eventually, around hour 30'000 the system switches to the anoxic state, with high abundance of both sulfur bacteria types. The oxygen diffusivity is then increased and at around hour 38'000 the system abruptly switches back to the oxic state.

@@

```{r uc1, echo=FALSE, fig.align="center", fig.cap='The temporal dynamics of the ecosystem model when an environmental condition (here parameter $a$, the oxygen diffusivity) changes. Plot of the steady states of the simulation runs under different oxygen diffusivity. In this simulation there is only one strain in each functional group. CB\textunderscore1 = cyanobacteria strain 1; SB\textunderscore1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show a figure adapted from the output of the $plot\\_dynamics()$  function.', message=FALSE, warning=FALSE, include=TRUE, messages=FALSE, out.width=350}
fig <- here::here("inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching.pdf")
if (!file.exists(fig)){
  library(ggplot2)
  library(microxanox)
  library(dplyr)
  x <- readRDS(here::here("inst/manuscript/data/hlh_with_defaults.RDS"))
  p <- plot_dynamics(x)
  p[[1]] <- p[[1]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[2]] <- p[[2]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[3]] <- p[[3]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p_a <- p[[4]]
  p[[4]]$data <- p[[4]]$data %>% dplyr::filter(species != "a")
  p[[4]]$data$species <- p[[4]]$data$species %>%
    dplyr::recode(SO = "Oxidised sulfur", O = "Oxygen", SR = "Sulfide", P = "Phosphorus")
  p[[4]] <- p[[4]] +
    xlab("time (h)") + 
    ylab(expression(atop(log[10](concentration),(µM)))) + 
    labs(colour = "substrate") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p_a$data <- p_a$data %>% dplyr::filter(species == "a")
  p_a$data$species <- p_a$data$species %>%
    dplyr::recode(a = "Oxygen diffusivity ")
  p_a <- p_a +
    ylab(expression(atop(Oxygen~diffusivity,(hr^{-1})))) + 
    labs(colour = "Environmental driver") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p <- p[[1]] / p[[2]] / p[[3]] / p[[4]] / p_a
  ggsave(
    filename = fig,
    plot = p,
    device = "pdf",
    width = 8,
    height = 8
  )  
}
knitr::include_graphics(fig)
```

@@

The package contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environmental change, because it is related to how difficult it can be to reserve the effects of environmental change [@Scheffer2001]. The amount of hysteresis is measured as the extent of the environmental condition (here oxygen diffusitivity) for which there were two steady states. I.e. it is the extent of the environmental conditions for which historical conditions play an important role in determining the current system state (a definition of hysteresis).

Using the package to calculate the extent of hysteresis involves setting ecosystem and simulation parameters, including parameters for the finding of steady states across an environmental gradient, running the steady state finding function, and analysing the results with the function that calculates extent of hysteresis. The code for this is provided in the [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html).

@@

Researchers wanting to model a new ecosystem do not, therefore, have to start from scratch. This will relieve researchers from needing to make software design decisions, and rather focus on appropriately representing their ecosystem and finding the results that interest them. Nevertheless, adaptation of the code in the package will require a person / persons that can take a conceptual model of an ecosystem and then represent that in terms of parameters and rate equations, and that is relatively proficient in R programming.

# Conclusions
<!-- > {>>Set out the conclusion of this original software publication.<<} -->

The *microxanox* R package allows the simulation, visualisation, and analysis of a model of a microbial ecosystem while allowing variation in the amount of diversity containing in each of the functional groups of organisms present. It has been used for the research described in another paper that describes one of the first investigations of the effects of diversity on ecosystem resilience @Limberger2022. In that paper, we show that diversity can have large and important effects of ecosystem responses, highlighting the need for models such as ours, with which one can easily manipulate the amount of biodiversity. The *microxanox* package has also been used to reproduce the results of the paper that inspired the package development [@Bush2017].

The package greatly lowers the amount of work required in further investigations of the specific ecosystem modelled. There has, for example, been quite limited investigation of how biodiversity influences the short-term responses of the modelled ecosystem to environmental change. Likewise, the package could be used to power an investigation of the effects of biodiversity on the usefulness of early warning signals of abrupt ecosystem change [@Scheffer2009]. In addition this package could be used as a template for the implementation for developing models of other types of ecosystems and organism. By doing so, other models can profit from the overall framework used, and the reproducibility aspects as well as the flexibility implemented.

>>>>

---
title: "Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states."
author:
- name: Rainer M Krug
  email: Rainer.Krug@uzh.ch;  Rainer@krugs.de
  affiliation: University of Zürich
  footnote: Corresponding Author
- name: Owen L. Petchey
  email: Owen.Petchey@ieu.uzh.ch
  affiliation: University of Zürich
address:
- code: University of Zürich
  address: Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse
    190, 8057 Zurich
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
keywords: 
  - reproducibility
  - regime shift 
  - steady state
  - final state
  - ordinary differential equations
abstract: |
  $Microxanox$ is an R package to simulate a three functional group ecosystem (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range.  The model can be implemented with different numbers of species per functional group. The package is constructed in such a way that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated. Furthermore, the package framework and code should serve as a useful starting point for making simulation models of other types of ecosystems.
journal: "SoftwareX"
date: "`r Sys.Date()`"

@@

<!-- # \textcolor{red}{TODO} -->

<!-- - [x] ~~**\textcolor{green}{update links}**~~ -->

<!-- - [x] ~~**\textcolor{green}{update cross references}**~~ -->

<!-- - [x] ~~**\textcolor{green}{add abstract}**~~ -->

<!-- - [x] ~~**\textcolor{green}{check keywords}**~~ -->

<!-- - [x] ~~**\textcolor{green}{add more references}**~~ -->

<!-- - [x] ~~**\textcolor{green}{expand on Impact section}**~~ -->

<!-- - [x] ~~**\textcolor{green}{expand on Conclusion section}**~~ -->

\pagebreak

# Required Metadata

## Current code version

Ancillary data table required for subversion of the codebase.

| **Nr.** | **Code metadata description**                                 | **Please fill in this column**                                                                                         |
|:------------:|:------------------------|:--------------------------------|
|   C1    | Current code version                                          | v0.9.1                                                                                                                 |
|   C2    | Permanent links to code/repository used for this code version | <https://github.com/UZH-PEG/microxanox>                                                                                |
|         |                                                               | [10.5281/zenodo.7148667](https://zenodo.org/record/7148667)                                                            |
|   C3    | Code Ocean compute capsule                                    |                                                                                                                        |
|   C4    | Legal Code License                                            | MIT                                                                                                                    |
|   C5    | Code versioning system used                                   | [git](https://git-scm.com)                                                                                             |
|   C6    | Software code languages, tools, and services used             | [R](https://cran.r-project.org/index.html)                                                                             |
|   C7    | Compilation requirements, operating environments              | [R (\>= 4.1.0)](https://cran.r-project.org/index.html)                                                                 |
|         |                                                               | magrittr                                                                                                               |
|         |                                                               | tibble                                                                                                                 |
|         |                                                               | ggplot2                                                                                                                |
|         |                                                               | patchwork                                                                                                              |
|         |                                                               | grDevices                                                                                                              |
|         |                                                               | stats                                                                                                                  |
|         |                                                               | mgcv                                                                                                                   |
|         |                                                               | deSolve                                                                                                                |
|         |                                                               | dplyr                                                                                                                  |
|         |                                                               | tidyr                                                                                                                  |
|         |                                                               | stringr                                                                                                                |
|         |                                                               | multidplyr                                                                                                             |
|   C8    | If available Link to developer documentation/manual           | [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html)                                       |
|   C9    | Support email for questions                                   | [Rainer.Krug\@uzh.ch](mailto:Rainer.Krug@uzh.ch){.email}; [Owen.Petchey\@ieu.uzh](mailto:Owen.Petchey@ieu.uzh){.email} |

@@

Mathematical models play a key role in the development of ecosystem models and in increasing the understanding of how and why ecosystems change when the environment changes [@Binzer2016a; @Chaparro-Pedraza2021; @Vasseur2005]. They are also important for developing hypotheses to test in empirical studies. One area of ecology in which models have been influential is the understanding of ecosystem responses to gradual change in an environmental driver [@Scheffer2001]. An environmental driver is an environmental condition that affects an ecosystem, but which is not itself affected by the ecosystem. An example driver would be the amount of nutrients entering a lake from run-off from surrounding agricultural land.

It is conceivable that an ecosystem state, such as the total biomass of a particular type of bacteria, may remain unchanged when an environmental driver changes. It is also possible that the ecosystem state changes gradually. It is also possible that the ecosystem state changes abruptly to a new state that is difficult to recover from [@Scheffer2001]. This possibility for abrupt, perhaps catastrophic changes, which are difficult to reverse, causes considerable concern [@Collins2021; @Northrop2021; @Vandermeer2019].

An example where a gradual change of an environmental variable causes an abrupt change in the system is the switch from an aerobic (oxygen is available for metabolism) to an anaerobic (oxygen generally unavailable) state in a microbial ecosystem. This system has been investigated by @Bush2017 in a simulation study of a mathematical model. Three types of microbes occur in the model: cyanobacteria (CB) dominating the oxic state, and two types of sulfur bacteria that dominate the anoxic state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria (PB)). The model shows that gradual change in the rate at which oxygen can diffuse into the ecosystem (termed oxygen diffusitivity) could cause catastrophic changes in the ecosystem state that would be difficult to reverse.

One feature of the study by @Bush2017 was limited biodiversity. Specifically, there was no biodiversity within each of the three types (i.e., functional groups) of bacteria. This leaves the question of if and how biodiversity within these functional groups might affect the ecosystem response to environmental change. This limitation is not specific to the study of @Bush2017. There are few if any studies of the effects of biodiversity on abrupt transitions between ecosystem states.

@@

# Software description

<!-- > {>>Describe the software in as much as is necessary to establish a vocabulary needed to explain its impact.<<} -->

@@

To make the simulation run with multiple species per functional group, we expressed different species' characteristics in the elements of vectors and matrices. We also coded the ordinary differential equations to include the vectors and matrices and used matrix mathematics. In this way, we made it possible to run simulations with different numbers of species without having to change the underlying code.

@@

## Software architecture

<!-- > {>>Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.<<} -->

The framework used when writing this package aimed to maximise simplicity for the user and to make it straightforward to reproduce results (see the supplement [@Petchey2022] to @Limberger2022 for an example of how this is used). As such, all the parameters needed to run a simulation or a set of simulations are contained in a single object (which can easily be created using included functions). This parameter object is given to a function that runs the simulations and returns the results. The returned results object is identical to the parameter object but with an additional slot named `results`, which contains the simulation results. Thus the returned results object contains the simulation conditions (parameters) as well as the results, and can be used to run the simulation again. This promotes reproducibility and makes incremental changes of individual parameters with a consecutive re-running of the simulations straightforward.

In the following sections we describe how to use the package to run one simulation and to find final states across an environmental gradient.

@@

## Examining ecosystem responses to environmental change

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's steady state (if it exists). The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity.

When one wishes to be able to make conclusions about how the steady state of the system is affected by the environmental driver, it is very important to note that the *final state* (found by the simulation) is not guaranteed to be a *steady state*. The user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper @Limberger2022, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow comparison of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study and was used to obtain the results in figures 3 and 4 of that article.

The second method runs two simulations, one with step-wise and slowly temporally *increasing* oxygen diffusivity, and the other with step-wise and slowly *decreasing* oxygen diffusivity. (Put another way, to explore the sensitivity of final states of the dynamic model under oxygen diffusivity variation, subset values of this parameter are numerically evaluated. Two value vectors of oxygen diffusivity are presented step-wise, gradually increasing and decreasing.) During this temporal environmental change, the state of the system is recorded just before change to a new oxygen diffusivity (we term this the *Temporal method*).

@@

From the results returned, summary measures about how the ecosystem final states changes with environmental change can be extracted. The function `get_stability_measures()` returns quantities such as the amount of environmental change required for the system to abruptly change to a different state.

@@

<!-- > {>>This is the main section of the article and the reviewers weight the description here appropriately -->

<!-- Indicate in what way new research questions can be pursued as a result of the software (if any). -->

<!-- Indicate in what way, and to what extent, the pursuit of existing research questions is improved (if so). -->

<!-- Indicate in what way the software has changed the daily practice of its users (if so). -->

<!-- Indicate how widespread the use of the software is within and outside the intended user group. -->

<!-- Indicate in what way the software is used in commercial settings and/or how it led to the creation of spin-off companies (if so).<<} -->

The open source implementation and extension of the model used in @Bush2017 provides the means of reproducing the results published while at the same time provides the means of making unique, innovative, and important investigations of how ecosystems change as the environmental changes, and how biodiversity may modulate this.

The design of the package code and functionality is with reproducibility in mind: the combination of all parameters being in a single parameter object and the return of the simulation as a result object which inherits from the parameter object, provides a relatively easy-to-use framework to implement reproducible experiments.

Here we evidence the impact of the *microxanox* package by describing three use-cases and then by describing how the package can be a starting point for models of other ecosystems. The first two use cases are described in detail (including the code for reproducing them) in the [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html) and the *Partial Reproduction* vignettes. The third is taken from @Limberger2022 and @Petchey2022.

@@

The study of @Bush2017 includes simulations of the effect of oxygen diffusivity (an environmental driver, in the sense that it affects the ecosystem but is not affected by it) on the ecosystem state (oxic or anoxic). The `microxonox` package contains functionality to make a specific temporal pattern of change in the oxygen diffusivity. As well as allowing individual simulations during which oxygen diffusivity varies, this functionality forms the basis of the temporal method for finding final states.

An example of this functionality is given in the *Partial Reproduction* vignette, which we briefly describe and show here (Figure \@ref(fig:uc1)). The example is composed of a single simulation, at the beginning of which the system is in the oxic state with high abundance of cyanobacteria. Oyxgen diffusivity is then slowly decreased and eventually, around hour 30'000 the system switches to the anoxic state, with high abundance of both sulfur bacteria types. The oxygen diffusivity is then increased and at around hour 38'000 the system abruptly switches back to the oxic state.

@@

```{r uc1, echo=FALSE, fig.align="center", fig.cap='The temporal dynamics of the ecosystem model when an environmental condition (here parameter $a$, the oxygen diffusivity) changes. Plot of the final states of the simulation runs under different oxygen diffusivity. In this simulation there is only one strain in each functional group. CB\textunderscore1 = cyanobacteria strain 1; SB\textunderscore1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show a figure adapted from the output of the $plot\\_dynamics()$  function.', message=FALSE, warning=FALSE, include=TRUE, messages=FALSE, out.width=350}
fig <- here::here("inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching.pdf")
if (!file.exists(fig)){
  library(ggplot2)
  library(microxanox)
  library(dplyr)
  x <- readRDS(here::here("inst/manuscript/data/hlh_with_defaults.RDS"))
  p <- plot_dynamics(x)
  p[[1]] <- p[[1]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[2]] <- p[[2]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[3]] <- p[[3]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p_a <- p[[4]]
  p[[4]]$data <- p[[4]]$data %>% dplyr::filter(species != "a")
  p[[4]]$data$species <- p[[4]]$data$species %>%
    dplyr::recode(SO = "Oxidised sulfur", O = "Oxygen", SR = "Sulfide", P = "Phosphorus")
  p[[4]] <- p[[4]] +
    xlab("time (h)") + 
    ylab(expression(atop(log[10](concentration),(µM)))) + 
    labs(colour = "substrate") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p_a$data <- p_a$data %>% dplyr::filter(species == "a")
  p_a$data$species <- p_a$data$species %>%
    dplyr::recode(a = "Oxygen diffusivity ")
  p_a <- p_a +
    ylab(expression(atop(Oxygen~diffusivity,(hr^{-1})))) + 
    labs(colour = "Environmental driver") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p <- p[[1]] / p[[2]] / p[[3]] / p[[4]] / p_a
  ggsave(
    filename = fig,
    plot = p,
    device = "pdf",
    width = 8,
    height = 8
  )  
}
knitr::include_graphics(fig)
```

@@

The package contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environmental change, because it is related to how difficult it can be to reserve the effects of environmental change [@Scheffer2001]. The amount of hysteresis is measured as the extent of the environmental condition (here oxygen diffusitivity) for which there were two steady states. I.e., it is the extent of the environmental conditions for which historical conditions play an important role in determining the current system state.

Using the package to calculate the extent of hysteresis involves setting ecosystem and simulation parameters, including parameters for the finding of final states across an environmental gradient, running the final state finding function, and analysing the results with the function that calculates extent of hysteresis. The code for this is provided in the [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html).

@@

Researchers wanting to model a new ecosystem do not, therefore, have to start from scratch. This will relieve researchers from needing to make software design decisions, and rather focus on appropriately representing their ecosystem and finding the results that interest them. Nevertheless, adaptation of the code in the package will require a person / persons that can take a conceptual model of an ecosystem and then represent that in terms of parameters and rate equations, and a person or persons  relatively proficient in R programming.

# Conclusions

<!-- > {>>Set out the conclusion of this original software publication.<<} -->

The *microxanox* R package allows the simulation, visualisation, and analysis of a model of a microbial ecosystem while allowing variation in the amount of diversity present in each of the functional groups of organisms present. It has been used for the research described in another paper that provides one of the first investigations of the effects of diversity on ecosystem resilience @Limberger2022. In that paper, we show that diversity can have large and important effects on ecosystem responses, highlighting the need for models such as ours, with which one can easily manipulate the amount of biodiversity. The *microxanox* package has also been used to reproduce the results of the paper that inspired the package development [@Bush2017].

The package greatly lowers the amount of work required in further investigations of the specific ecosystem modelled. There has, for example, been quite limited investigation of how biodiversity influences the short-term responses of the modelled ecosystem to environmental change. Likewise, the package could be used to power an investigation of the effects of biodiversity on the usefulness of early warning signals of abrupt ecosystem change [@Scheffer2009]. In addition this package could be used as a template for developing models of other types of ecosystems and organism. By doing so, other models can profit from the overall framework used, and the reproducibility aspects as well as the flexibility implemented.

# Conflict of Interest
